Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-20 tex/manuscript.tex
Kun mindre rettelser

Solving $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
\begin{equation}
X_0^t(X_0X_0^t)^{-1} \begin{bmatrix}B_0\\0\end{bmatrix}
=[B_0^t|C_0^t] \begin{bmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{bmatrix} B_0
=\rho_{C_0}B_0^tPB_0
\end{equation}
with $P=(B_0\rho_{C_0}B_0^t)^{-1}$, and $\rho_{W}=I-W^t(WW^t)^{-1}W$ for any $k\times n$ matrix with $k\le n$ defining the projection to the $n-k$-hyperplane perpendicular to the $k$-hyperplane spanned by $W$, yields
\begin{equation}
(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L)
\quad\text{where}\quad
L=B_0^tPB_0
\end{equation}
since $\epsilon_0\sim\text{N}(0,\sigma^2I)$. Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L$: this holds since $L$ is a symmetric rank $q$ matrix,  has the same eigenvalues as $PB_0B_0^t$ the inverse of which, $(B_0B_0^t)^{-1}B_0\rho_{C}B_0^t$, has eigenvalues in $(0,1)$ (not 0 since $B_0$ and $C_0$ are linearly independent). Then,
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2=\sigma^2\sum_{i=1}^q\lambda_iU_i
\quad\text{where}\quad
U_1,\ldots,U_q\sim\chi^2_1.
\end{equation}
Approximating this so that the first two momenta match gives
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2\approx\tilde\sigma^2\chi^2_{\tilde q}
\quad\text{where}\quad
\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L),\,
\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L^2),
\end{equation}
from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are linearly independent, i.e. $B_0C_0^t=0$.

We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices. The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0|U']$ so that $AU_0=0$ while $AU'$ is non-singular. Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$. This allows us to express
\begin{equation}
L=U_0^tB^tPBU_0
\quad\text{where}\quad
P=(BU_0\rho_{C_0}U_0^tB^t)^{-1}=(B\rho_{A,C}B^t)^{-1}
\end{equation}
and $\rho_{A,C}=\rho_{[A^t|C^t]^t}$ is the projection to the $n-p-r$-hyperplane perpendicular to the $p+r$-hyperplane spanned by $A$ and $C$. We can then compute
\begin{equation}
\text{tr}(L)=\text{tr}(M),\,
\text{tr}(L^2)=\text{tr}(M^2)
\quad\text{where}\quad
M=BU_0U_0^tB^tP=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
\end{equation}
and where $M$ and $L$ have the same eigenvalues.
Notice that if $B$ is orthogonal to $A$, which can be achieved by replacing $B$ with $B-VA$ for some matrix $V$, then $B\rho_A=B$. If $C$ is also made orthogonal to $A$, then $\rho_{A,C}$ can be replaced by $\rho_C$.

>>>>

Solving $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
\begin{equation}
X_0^t(X_0X_0^t)^{-1} \begin{bmatrix}B_0\\0\end{bmatrix}
=[B_0^t|C_0^t] \begin{bmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{bmatrix} B_0
=\rho_{C_0}B_0^tPB_0
\end{equation}
with $P=(B_0\rho_{C_0}B_0^t)^{-1}$, and $\rho_{W}=I-W^t(WW^t)^{-1}W$ for any $k\times n$ matrix with $k\le n$ defining the projection to the $n-k$-hyperplane perpendicular to the $k$-hyperplane spanned by $W$, yields
\begin{equation}
(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L)
\quad\text{where}\quad
L=B_0^tPB_0
\end{equation}
since $\epsilon_0\sim\text{N}(0,\sigma^2I)$. Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L$: this holds since $L$ is a symmetric rank $q$ matrix,  has the same eigenvalues as $PB_0B_0^t$ the inverse of which, $(B_0B_0^t)^{-1}B_0\rho_{C}B_0^t$, has eigenvalues in $(0,1)$ (not 0 since $B_0$ and $C_0$ are linearly independent). Then,
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2=\sigma^2\sum_{i=1}^q\lambda_iU_i
\quad\text{where}\quad
U_1,\ldots,U_q\sim\chi^2_1.
\end{equation}
Approximating this so that the first two momenta match gives
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2\approx\tilde\sigma^2\chi^2_{\tilde q}
\quad\text{where}\quad
\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L),\,
\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L^2),
\end{equation}
from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, and where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are orthogonal, i.e. $B_0C_0^t=0$.

We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices. The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0|U']$ so that $AU_0=0$ while $AU'$ is non-singular. Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$. This allows us to express
\begin{equation}
L=U_0^tB^tPBU_0
\quad\text{where}\quad
P=(BU_0\rho_{C_0}U_0^tB^t)^{-1}=(B\rho_{A,C}B^t)^{-1}
\end{equation}
and $\rho_{A,C}=\rho_{[A^t|C^t]^t}$ is the projection to the $n-p-r$-hyperplane perpendicular to the $p+r$-hyperplane spanned by $A$ and $C$. We can then compute
\begin{equation}
\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\text{tr}(L)=\text{tr}(M),\,
\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\text{tr}(L^2)=\text{tr}(M^2)
\quad\text{with}\quad
M=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
\end{equation}
since $M=BU_0U_0^tB^tP$ and $L=U_0^tB^tPBU_0$ have the same non-zero eigenvalues.
Notice that if $B$ is orthogonal to $A$, which can be achieved by replacing $B$ with $B-VA$ for some matrix $V$, then $B\rho_A=B$. If $C$ is also made orthogonal to $A$, then $\rho_{A,C}$ can be replaced by $\rho_C$.
